---
title: "Getting started with dbpmr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dbpmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5
)
# The model writes output files to disk, so run all example chunks inside a
# temporary directory rather than the user's working directory.
knitr::opts_knit$set(root.dir = tempdir())
```

`dbpmr` implements the **Dynamic Benthic Pelagic Model (DBPM)**: a coupled,
size-structured model that links a pelagic predator size spectrum to a benthic
detritivore spectrum and a detritus pool, integrated forward in time by a C
simulation engine. The model is based on Blanchard et al. (2009, 2011), with the
spatial movement of the pelagic spectrum from Castle et al. (2011); see
`citation("dbpmr")`.

This vignette walks through a minimal, aspatial (0-D) run.

```{r library}
library(dbpmr)
```

## 1. Configure the run and the grid

`Setup.Run()` defines the high-level structure of the simulation (how many
species, whether benthic and pelagic are coupled, the spatial dimension) and
creates an output directory named after the run. `Setup.Grid()` defines the
log-mass and time discretisation.

```{r run-grid}
run  <- Setup.Run("MyRun", no_pelagic = 1, no_benthic = 1,
                  spatial_dim = 0, coupled_flag = TRUE, diff_method = 1)
grid <- Setup.Grid(run, tmax = 10)   # integrate for 10 years

run
```

## 2. Configure the resource, species and detritus

Each component is built with its own helper, which returns a parameter object
and creates the component's output sub-directory. Here we use a single pelagic
species and a single benthic species.

By default (`rep_method = 2`), reproduction is allocated from the assimilated
energy of mature individuals, so the run is self-contained and needs no extra
input files. Other options are a fixed recruitment (`0`), a prescribed
reproduction time series read from a file (`1`, prepared with `Setup.Rep()`),
and reproduction proportional to mature biomass (`3`); see `?Setup.Pelagic`.

> **Note:** `rep_method = 1` requires a reproduction time-series input file.
> These data are **not** provided with the package and must be supplied by the
> user (via `Setup.Rep()`); otherwise `SizeSpectrum()` stops with an error.

```{r species}
plankton <- Setup.Plankton(run, filename = "plankton")
pelagic  <- Setup.Pelagic(run,  filename = "fish")
benthic  <- Setup.Benthic(run,  filename = "benthos")
detritus <- Setup.Detritus(run, filename = "detritus")
```

## 3. Run the simulation

`SizeSpectrum()` assembles the parameters, calls the C engine and writes the
output text files under `MyRun/`. It returns the paths needed to read the
results back in. (The engine prints a progress counter, which we suppress here.)

```{r run-model}
invisible(capture.output(
  files <- SizeSpectrum(run, grid, plankton, pelagic, benthic, detritus)
))
```

## 4. Read the results back

`Read.In()` reads one component's output and returns a results object whose
class depends on the component type.

```{r read}
fish <- Read.In("MyRun", "fish")
class(fish)
```

## 5. Plot the size spectrum

`Extract.Time()` pulls the spectrum at a single output time, and
`Plot.Spectrum()` plots it (here, log abundance against log body mass at the end
of the run).

```{r spectrum}
final <- Extract.Time(fish, time = max(fish@trange))
Plot.Spectrum(final, type = "l",
              xlab = "log body mass", ylab = "log abundance",
              main = "Pelagic size spectrum (final time step)")
```

We can also look at how total pelagic biomass changes through time, using the
summary values stored on the results object:

```{r biomass}
plot(fish@trange, fish@biomass, type = "l",
     xlab = "Time (years)", ylab = "Pelagic biomass")
```

## Where to go next

- `Average.Time()` returns the spectrum averaged over a time window, and
  `Animate()` steps through the output times.
- `Setup.ts()`, `Setup.Rep()` and `Setup.fishing()` add initial conditions,
  reproduction time series and fishing mortality.
- Set `spatial_dim = 1` or `2` in `Setup.Run()` for spatial runs (pelagic
  movement only).

See the function help pages (e.g. `?Setup.Pelagic`, `?SizeSpectrum`) for the
full set of parameters.

## References

- Blanchard, J.L., Jennings, S., Law, R., Castle, M.D., McCloghrie, P.,
  Rochet, M.-J. & Benoît, E. (2009). How does abundance scale with body size in
  coupled size-structured food webs? *Journal of Animal Ecology*, **78**(1),
  270–280.
- Blanchard, J.L., Law, R., Castle, M.D. & Jennings, S. (2011). Coupled energy
  pathways and the resilience of size-structured food webs. *Theoretical
  Ecology*, **4**(3), 289–300.
- Castle, M.D., Blanchard, J.L. & Jennings, S. (2011). Predicted effects of
  behavioural movement and passive transport on individual growth and community
  size structure in marine ecosystems. *Advances in Ecological Research*,
  **45**, 41–66.
