---
title: "Getting started with dbpmr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dbpmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5
)
# The model writes output files to disk, so run all example chunks inside a
# temporary directory rather than the user's working directory.
knitr::opts_knit$set(root.dir = tempdir())
```

`dbpmr` implements the **Dynamic Benthic Pelagic Model (DBPM)**: a coupled,
size-structured model that links a pelagic predator size spectrum to a benthic
detritivore spectrum and a detritus pool, integrated forward in time by a C
simulation engine. The model is based on Blanchard et al. (2009, 2011), with the
spatial movement of the pelagic spectrum from Castle et al. (2011); see
`citation("dbpmr")`.

This vignette walks through a minimal, aspatial (0-D) run.

```{r library}
library(dbpmr)
```

## 1. Configure the run and the grid

`Setup.Run()` defines the high-level structure of the simulation (how many
species, whether benthic and pelagic are coupled, the spatial dimension) and
creates an output directory named after the run. `Setup.Grid()` defines the
log-mass and time discretisation.

```{r run-grid}
run  <- Setup.Run("MyRun", no_pelagic = 1, no_benthic = 1,
                  spatial_dim = 0, coupled_flag = TRUE, diff_method = 1)
grid <- Setup.Grid(run, tmax = 20)   # integrate for 20 years

run
```

### Customising the grid

`Setup.Grid()` has sensible defaults, so for a simple run you only need to set
the few arguments you care about — usually how long to run and how often to
save output:

```r
grid <- Setup.Grid(run,
  # mass grid (log body mass)
  mstep    = 0.1 * log(10),  # computation step (0.1 log10 unit)
  moutstep = 0.1 * log(10),  # output step (must be a multiple of mstep)
  # time grid (years)
  tmax     = 20,             # run length
  tstep    = 1/365,          # integration step (daily)
  toutstep = 73/365          # output step (must be a multiple of tstep)
)
```

The mass grid defaults to the canonical FishMIP/DBPM log10 spectrum (`10^-12` to
`10^6` g in steps of `0.1` log10 unit, consumer minimum `10^-3` g), which already
spans the default plankton and consumer ranges, so you rarely need to set
`mmin`/`mmax`. For a `spatial_dim = 0` run the spatial arguments are ignored and
the x/y dimensions are collapsed to a single cell.

**The one rule that trips people up:** the *output* steps must be exact
whole-number multiples of the *computation* steps, or `SizeSpectrum()` stops
with an error. With the default daily `tstep = 1/365`:

| Desired output | `toutstep` | = whole steps? |
|---|---|---|
| every ~73 days (default) | `73/365` | 73 ✓ |
| weekly | `7/365` | 7 ✓ |
| ~monthly | `30/365` | 30 ✓ |
| annual | `1` | 365 ✓ |
| calendar-monthly | `1/12` | 30.4 ✗ (errors) |

The same multiple-of rule applies to `moutstep` vs `mstep`, and to
`xoutstep`/`xstep` and `youtstep`/`ystep` for spatial runs.

## 2. Configure the resource, species and detritus

Each component is built with its own helper, which returns a parameter object
and creates the component's output sub-directory. Here we use a single pelagic
species and a single benthic species.

By default (`rep_method = 2`), reproduction is allocated from the assimilated
energy of mature individuals, so the run is self-contained and needs no extra
input files. Other options are a fixed recruitment (`0`), a prescribed
reproduction time series read from a file (`1`, prepared with `Setup.Rep()`),
and reproduction proportional to mature biomass (`3`); see `?Setup.Pelagic`.

> **Note:** `rep_method = 1` requires a reproduction time-series input file.
> These data are **not** provided with the package and must be supplied by the
> user (via `Setup.Rep()`); otherwise `SizeSpectrum()` stops with an error.

```{r species}
plankton <- Setup.Plankton(run, filename = "plankton")
pelagic  <- Setup.Pelagic(run,  filename = "fish")
benthic  <- Setup.Benthic(run,  filename = "benthos")
detritus <- Setup.Detritus(run, filename = "detritus")
```

## 3. Run the simulation

`SizeSpectrum()` assembles the parameters, calls the C engine and writes the
output text files under `MyRun/`. It returns the paths needed to read the
results back in. (The engine prints a progress counter, which we suppress here.)

```{r run-model}
invisible(capture.output(
  files <- SizeSpectrum(run, grid, plankton, pelagic, benthic, detritus)
))
```

## 4. Read the results back

`Read.In()` reads one component's output and returns a results object whose
class depends on the component type. Here we read both the pelagic and the
benthic results.

```{r read}
fish    <- Read.In("MyRun", "fish")
benthos <- Read.In("MyRun", "benthos")
class(fish)
```

## 5. Plot the coupled size spectrum

`Extract.Time()` pulls the spectrum at a single output time. `Plot.Spectrum()`
opens the plot and `Points.Spectrum()` adds a second spectrum on the same axes,
so we can show the pelagic and benthic spectra together at the end of the run.
The benthos sit at smaller body sizes than the fish.

```{r spectrum}
fish_final    <- Extract.Time(fish,    time = max(fish@trange))
benthos_final <- Extract.Time(benthos, time = max(benthos@trange))

Plot.Spectrum(fish_final, type = "l", col = "steelblue", lwd = 2,
              ylim = c(-25, 2),
              xlab = "log body mass", ylab = "log abundance",
              main = "Coupled size spectrum (final time step)")
Points.Spectrum(benthos_final, col = "darkorange", pch = 20)
legend("topright", legend = c("pelagic (fish)", "benthic"),
       col = c("steelblue", "darkorange"),
       lty = c(1, NA), pch = c(NA, 20), lwd = c(2, NA), bty = "n")
```

We can also track how the total biomass of each component changes through time,
using the summary values stored on the results objects. Plotting on a
logarithmic y-axis keeps the early transient from dominating the view: the
benthic biomass settles to a steady state within a few years, while the pelagic
biomass declines and roughly levels off.

```{r biomass}
plot(fish@trange, fish@biomass, type = "l", col = "steelblue", lwd = 2,
     log = "y", ylim = range(c(fish@biomass, benthos@biomass)),
     xlab = "Time (years)", ylab = "Biomass (log scale)",
     main = "Biomass through time")
lines(benthos@trange, benthos@biomass, col = "darkorange", lwd = 2)
legend("right", legend = c("pelagic (fish)", "benthic"),
       col = c("steelblue", "darkorange"), lwd = 2, bty = "n")
```

This minimal default configuration is illustrative rather than tuned: the
populations are not started from a steady state, so the pelagic spectrum decays
over the run. Realistic applications set initial conditions, reproduction and
fishing inputs (next section) and choose parameters that give a stable spectrum.

## Where to go next

- `Average.Time()` returns the spectrum averaged over a time window, and
  `Animate()` steps through the output times.
- `Setup.ts()`, `Setup.Rep()` and `Setup.fishing()` add initial conditions,
  reproduction time series and fishing mortality.
- Set `spatial_dim = 1` or `2` in `Setup.Run()` for spatial runs (pelagic
  movement only).

See the function help pages (e.g. `?Setup.Pelagic`, `?SizeSpectrum`) for the
full set of parameters.

## Driving the model with spatial data

For a spatial run (`spatial_dim = 1` or `2`) the model lives on a **regular
rectangular x/y grid** and reads its inputs as plain text. It does **not** read
shapefiles, rasters or netCDF directly, and it has no land mask or environmental
forcing of vital rates. You can, however, use R's geospatial packages to sample
such data onto the model grid and feed the inputs that *are* spatial: initial
conditions / time series (`Setup.ts()`) and fishing mortality (`Setup.fishing()`).

The grid cells are the Cartesian product of

```r
xs <- seq(grid@xmin, grid@xmax, grid@xstep)
ys <- seq(grid@ymin, grid@ymax, grid@ystep)
ts <- seq(0, grid@tmax, grid@tstep)
```

and the input files order rows as **time outermost, then x, then y innermost** —
i.e. the row index is `(t-1) * length(xs) * length(ys) + (x-1) * length(ys) + y`.

**Example — spatially varying fishing mortality from a raster.** `Setup.fishing()`
accepts a function of `(m, t, x, y)`, evaluated over the whole grid, so you can
wrap a raster lookup directly (here fishing is size- and time-independent, so `m`
and `t` are ignored):

```r
library(terra)

# Your raster, already in the model's coordinate units/extent (project first if
# it is in lon/lat — the model uses plain numeric x/y, not geographic coords).
Fmap <- rast("fishing_mortality.tif")

Ffun <- function(m, t, x, y) terra::extract(Fmap, cbind(x, y))[1, 1]

run     <- Setup.Run("Spatial", 1, 0, spatial_dim = 2, TRUE, 1)
grid    <- Setup.Grid(run, xmax = 1000, xstep = 50, ymax = 1000, ystep = 50, tmax = 10)
pelagic <- Setup.Pelagic(run, filename = "fish", fishing_flag = TRUE)

Setup.fishing(pelagic, run, grid, func = Ffun)   # writes Spatial/Input/fish_fishing_ts.txt
```

For a **netCDF** time series (one layer per time slice), pick the layer for the
model time `t`:

```r
Fnc <- rast("fishing_mortality.nc")              # terra reads netCDF too
layer_for <- function(t) ...                     # map model time t -> layer index
Ffun <- function(m, t, x, y) terra::extract(Fnc[[layer_for(t)]], cbind(x, y))[1, 1]
```

If you would rather build the field explicitly (and faster than a per-cell
lookup), construct the matrix in the row order above and pass it to
`Setup.ts(..., mat = ...)` for a spatio-temporal initial/time-series input:

```r
cells <- expand.grid(y = ys, x = xs, t = ts)     # y fastest, then x, then t
# fill one column per mass bin of the species; rows follow `cells`
```

**Shapefiles** (`sf`) work the same way once converted to per-cell values — e.g.
point-in-polygon to tag each grid cell with a region, or `terra::rasterize()` a
polygon attribute onto `xs`/`ys`. Remember the model still simulates *every*
rectangular cell: a shapefile lets you assign spatially varying values, but it
cannot mask out land or give the domain an irregular shape.

## References

- Blanchard, J.L., Jennings, S., Law, R., Castle, M.D., McCloghrie, P.,
  Rochet, M.-J. & Benoît, E. (2009). How does abundance scale with body size in
  coupled size-structured food webs? *Journal of Animal Ecology*, **78**(1),
  270–280.
- Blanchard, J.L., Law, R., Castle, M.D. & Jennings, S. (2011). Coupled energy
  pathways and the resilience of size-structured food webs. *Theoretical
  Ecology*, **4**(3), 289–300.
- Castle, M.D., Blanchard, J.L. & Jennings, S. (2011). Predicted effects of
  behavioural movement and passive transport on individual growth and community
  size structure in marine ecosystems. *Advances in Ecological Research*,
  **45**, 41–66.
