---
title: "Running the gridded model for a region using FishMIP inputs (Kerguelen)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the gridded model for a region using FishMIP inputs (Kerguelen)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5,
                      fig.align = "center")
```

This vignette shows how to drive `dbpmr` **cell by cell over a region**, forced by
FishMIP Earth System Model (ESM) inputs. The worked example is the **Kerguelen
Plateau** (a sub-Antarctic FishMIP regional-model domain in the Indian sector of
the Southern Ocean). Each 1° cell runs an independent coupled benthic–pelagic
size-spectrum model driven by its own local plankton spectrum and temperature.

To keep the example self-contained and laptop-friendly it is **unfished**: it maps
the equilibrium biomass the ESM forcing supports, with no fishing or calibration.
Adding fishing effort and calibrating catchability is a further step — see
*Going further* at the end.

## 1. Where the inputs come from

Two ingredients define a regional run:

* **The region boundary.** FishMIP publishes a shapefile of all regional-model
  domains at [Fish-MIP/FishMIP_regions](https://github.com/Fish-MIP/FishMIP_regions)
  (`outputs/FishMIP_regional_models.zip`). We used the **"Kerguelen Mizer"**
  polygon and kept the 1° cell centres inside it — **84 cells**.
* **The ESM forcing.** Per-region gridded forcing (temperature, phytoplankton
  carbon, export, …) can be downloaded from the
  [FishMIP Regional Climate Forcing Data Explorer](https://fishmip.global-ecosystem-model.cloud.edu.au/shiny/FishMIP_Input_Explorer/)
  (gridded `.zarr` for 3-D variables, tabular `.parquet` for summaries), or taken
  directly from the ISIMIP3a GFDL-MOM6-COBALT2 obsclim outputs.

From those raw fields we derive, **per cell and per month**, the two things the
engine needs as plankton forcing — the size-spectrum **intercept** and **slope** —
plus the **experienced** (vertically biomass-weighted) water-column temperature,
the sea-floor temperature, and the export ratio. That processing is summarised
below (not run here; it needs the full ESM netCDFs) and is described in full in
the package's gridded workflow:

```{r eval = FALSE}
# Sketch of the offline forcing build (per cell, per month), from the ESM fields:
#   intercept, slope <- GetPPIntSlope(<phyc>, <phypico>)   # <.> = vertical biomass-weighted mean, 0-200 m
#   texp             <- <thetao>                            # phyc-weighted experienced temperature
#   tob              <- sea-floor temperature
#   export           <- expc-bot / intpp
# The result is one row per (cell, year, month) — exactly what we bundle below.
```

The processed forcing for Kerguelen ships with the package so this vignette runs
without any downloads.

## 2. Load the bundled example inputs

```{r}
library(dbpmr)

ex     <- function(f) system.file("extdata", f, package = "dbpmr")
cells  <- read.csv(ex("kerguelen_cells.csv"))                 # lon, lat, depth (m)
forc   <- readRDS(ex("kerguelen_forcing.rds"))                # per cell x month forcing
params <- jsonlite::fromJSON(ex("kerguelen_biology_params.json"))$params  # global biology constants

nrow(cells)                                                   # 84 cells
range(forc$year)                                              # 1961-2010, monthly
head(forc, 3)
```

Each cell's forcing is a monthly series; we reshape it into `months × cells`
matrices for convenient indexing.

```{r}
ckey <- paste(round(forc$lon), round(forc$lat))
cid  <- match(paste(round(cells$lon), round(cells$lat)), unique(ckey))
nmo  <- nrow(forc) / nrow(cells)
mat  <- function(v) matrix(forc[[v]], nmo, nrow(cells))
Pint <- mat("intercept"); Pslp <- mat("slope"); Ptex <- mat("texp")
Ptob <- mat("tob");       Pexp <- mat("export")
yr1  <- 1:12                                                   # first-year month columns
```

## 3. A single-cell run

The helper below runs **one cell**, unfished, for `nyr` years, driven by that
cell's 12 monthly forcing values. It is a thin wrapper around the standard
`Setup.*` → `SizeSpectrum()` calls: a plankton spectrum whose intercept/slope come
from the forcing (`ts_flag = TRUE`), a temperature/`export` forcing table, and one
pelagic + one benthic size class. Passing a `state` warm-restarts from a previous
year.

```{r}
LN10 <- log(10)
Sys.setenv(PEL_IMM_FRAC = "0.15")        # connectivity/recruitment floor (engine env var)
ascl <- 1/3; mu0s <- 0.5                 # search-volume scaling A/3, natural-mortality scaling
dh <- params$defecate_prop; dl <- params$def_low
Kp <- (1-dh)*params$growth_pred; Rp <- (1-dh)*(1-(params$growth_pred+params$energy_pred)); Ep <- (1-dh)*params$energy_pred
Kl <- (1-dl)*params$growth_detritivore; Rl <- (1-dl)*(1-(params$growth_detritivore+params$energy_detritivore)); El <- (1-dl)*params$energy_detritivore
te <- function(T) exp(params$c1 - params$activation_energy/(params$boltzmann*(T+273)))

run_cell <- function(depth, cint, cslp, ctex, ctob, cexp, nyr, state = NULL) {
  wd <- tempfile("cell"); dir.create(wd); old <- setwd(wd)
  on.exit({ setwd(old); unlink(wd, recursive = TRUE, force = TRUE) }, add = TRUE)
  dcorr <- min(depth, 200); prefben <- 0.8 * exp(-depth/1500)
  mon   <- function(t) (floor(t*12) %% 12) + 1
  plfun <- function(m, t, x, y) { j <- mon(t); (10^cint[j]/LN10) * exp(cslp[j]*m) }
  run  <- Setup.Run("R", 1, 1, 0, TRUE, 1)
  grid <- Setup.Grid(run, tmax = nyr, tstep = 1/48, toutstep = 1)
  pl   <- Setup.Plankton(run, filename = "plankton", lambda = cslp[1], ts_flag = TRUE)
  Setup.ts(pl, run, grid, func = plfun)
  pe <- Setup.Pelagic(run, filename = "fish", mmin = -3*LN10, mmat = 2*LN10, mmax = 6*LN10,
                      alpha = params$metabolic_req_pred, A = 64*ascl, mu_0 = params$natural_mort*mu0s,
                      pref_ben = prefben, K_pla = Kp, R_pla = Rp, Ex_pla = Ep, K_pel = Kp, R_pel = Rp,
                      Ex_pel = Ep, K_ben = Kl, R_ben = Rl, Ex_ben = El, rep_method = 2, fishing_flag = FALSE)
  be <- Setup.Benthic(run, filename = "benthos", mmin = -3*LN10, mmat = 1*LN10, mmax = 4*LN10,
                      alpha = params$metabolic_req_detritivore, A = 6.4*ascl, mu_0 = params$natural_mort*mu0s,
                      K_det = Kl, R_det = Rl, Ex_det = El, rep_method = 2, fishing_flag = FALSE)
  de <- Setup.Detritus(run, filename = "detritus")
  din <- file.path("R", "Input"); dir.create(din, showWarnings = FALSE, recursive = TRUE)
  nst <- round(nyr/(1/48)) + 2; jm <- ((floor(seq_len(nst)*(1/48)*12)) %% 12) + 1
  writeLines(c("pel_tempeff,ben_tempeff,sinking_rate,depth",
               sprintf("%.8g,%.8g,%.8g,%.8g", te(ctex[jm]), te(ctob[jm]), pmax(cexp[jm], 0), dcorr)),
             file.path(din, "forcing_ts.txt"))
  if (!is.null(state)) { pe@initial_flag <- TRUE; be@initial_flag <- TRUE; de@initial_flag <- TRUE
    writeLines(paste(state$U, collapse = ","), file.path(din, "fish_ts.txt"))
    writeLines(paste(state$V, collapse = ","), file.path(din, "benthos_ts.txt"))
    writeLines(paste(state$W, collapse = ","), file.path(din, "detritus_ts.txt")) }
  invisible(utils::capture.output(SizeSpectrum(run, grid, pl, pe, be, de)))
  f <- Read.In("R", "fish"); b <- Read.In("R", "benthos"); d <- Read.In("R", "detritus")
  m <- f@mrange; dm <- diff(m)[1]
  U <- as.numeric(f@finaluvals[1, -(1:3)]); V <- as.numeric(b@finaluvals[1, -(1:3)]); W <- tail(as.numeric(d@biomass), 1)
  list(state = list(U = U, V = V, W = W), m = m, U = U, V = V,
       BU = sum(U*exp(m)*dm)*dcorr, BV = sum(V*exp(m)*dm)*20)   # depth-integrated pelagic / benthic biomass
}
```

Let's spin a handful of cells (a north–south transect) to equilibrium and look at
the pelagic and benthic biomass they support:

```{r}
sel <- cells[order(cells$lat), ][round(seq(1, nrow(cells), length.out = 5)), ]
demo <- lapply(seq_len(nrow(sel)), function(i) {
  co <- which(round(cells$lon) == round(sel$lon[i]) & round(cells$lat) == round(sel$lat[i]))[1]
  run_cell(cells$depth[co], Pint[yr1, co], Pslp[yr1, co], Ptex[yr1, co], Ptob[yr1, co], Pexp[yr1, co],
           nyr = 25)
})
data.frame(lon = sel$lon, lat = sel$lat, depth = round(sel$depth),
           pelagic_B = round(sapply(demo, `[[`, "BU"), 1),
           benthic_B = round(sapply(demo, `[[`, "BV"), 1))
```

Running all 84 cells and stepping them through the full 1961–2010 monthly forcing
(warm-restarting each year) is the same call in a loop — it takes a few minutes on
a laptop. We ship the finished output so the maps below render instantly.

## 4. Region result

```{r fig.height = 4}
res <- readRDS(ex("kerguelen_run_result.rds"))
ce  <- res$cells

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
draw <- function(z, ttl) {
  pal <- hcl.colors(64, "YlGnBu", rev = TRUE)
  plot(ce$lon, ce$lat, pch = 15, cex = 2.2, col = pal[as.integer(cut(z, 64))],
       xlab = "Longitude", ylab = "Latitude", main = ttl)
  points(c(69.2, 73.5), c(-49.3, -53.1), pch = 17)            # Kerguelen & Heard Is.
}
draw(ce$BU_recent, "Pelagic biomass")
draw(ce$BV_recent, "Benthic biomass")
```

Pelagic biomass peaks over the **northern plateau** near the islands — where the
ESM forcing carries the most phytoplankton — and declines into the colder, less
productive south.

```{r fig.height = 4}
par(mfrow = c(1, 1), mar = c(4, 4, 2, 1))
matplot(res$yrs, cbind(colMeans(res$BU, na.rm = TRUE), colMeans(res$BV, na.rm = TRUE)),
        type = "l", lty = 1, lwd = 2, col = c("#1f78b4", "#b2570b"),
        xlab = "Year", ylab = expression("Regional mean biomass (g m"^-2*")"))
legend("right", c("Pelagic predators", "Benthic detritivores"),
       col = c("#1f78b4", "#b2570b"), lwd = 2, bty = "n")
```

And the size spectra the model produces in a productive (northern) versus a
cooler (southern) cell:

```{r fig.height = 4}
ok <- which(vapply(res$spec, is.list, logical(1)))
iN <- ok[which.max(ce$lat[ok])]; iS <- ok[which.min(ce$lat[ok])]
sN <- res$spec[[iN]]; sS <- res$spec[[iS]]
plot(sN$m/LN10, log10(pmax(sN$U, 1e-30)), type = "l", lwd = 2, col = "#1f78b4", ylim = c(-30, 5),
     xlab = expression(log[10]~"body mass (g)"), ylab = expression(log[10]~"abundance density"))
lines(sN$m/LN10, log10(pmax(sN$V, 1e-30)), lwd = 2, col = "#b2570b")
lines(sS$m/LN10, log10(pmax(sS$U, 1e-30)), lwd = 2, lty = 2, col = "#1f78b4")
lines(sS$m/LN10, log10(pmax(sS$V, 1e-30)), lwd = 2, lty = 2, col = "#b2570b")
legend("topright", c("pelagic (north)", "benthic (north)", "pelagic (south)", "benthic (south)"),
       col = c("#1f78b4", "#b2570b", "#1f78b4", "#b2570b"), lty = c(1, 1, 2, 2), lwd = 2, bty = "n", cex = 0.85)
```

## 5. Adding fishing and calibrating to catch

The run above is unfished. To make it a *calibrated* regional model we add two more
FishMIP inputs — both downloadable per region from the
[FishMIP Input Explorer](https://fishmip.global-ecosystem-model.cloud.edu.au/shiny/FishMIP_Input_Explorer/):

* **fishing effort** (nominal effort by year), which drives fishing pressure, and
* **observed catch** (by functional group), the calibration target.

Effort is redistributed across cells each year by a **spatial-gravity** model
(more effort where accessible, high-biomass cells are), fishing is applied to size
classes above ~30 cm (separately for pelagic vs benthic/demersal groups mapped from
the catch functional groups), and the two catchabilities `q_pel`, `q_ben` are fitted
so aggregated modelled catch matches the observed series. Because Kerguelen has no
0-D anchor, we calibrate **directly on the grid** (spin cached once, then BOBYQA on
`log10(q)` with each objective evaluation running the fished transient). That step is
slow (many transients), so it is shown here but not run — we load its shipped result:

```{r eval = FALSE}
# Gridded-native q calibration (sketch; minutes-to-hours depending on region size):
#   spin unfished (cache) -> for each candidate (q_pel, q_ben):
#     transient 1961-2010 with annual gravity redistribution of FishMIP effort,
#     aggregate per-cell catch to a regional density series,
#     objective = wlogmse(agg_pel, obs_pel) + wlogmse(agg_ben, obs_ben)
#   BOBYQA over log10(q); observed catch split pelagic/benthic by functional group.
```

```{r fig.height = 4}
cal <- readRDS(ex("kerguelen_calib.rds"))
c(q_pel = cal$qg[1], q_ben = cal$qg[2])                       # fitted catchabilities
c(cor_pelagic = cal$corr_pel, cor_benthic = cal$corr_ben)    # modelled-vs-observed catch correlation

par(mfrow = c(1, 2), mar = c(4, 4.2, 3, 1))
y <- cal$yrs; pos <- function(x) pmax(x, 1e-9)
yl <- range(c(cal$pel, cal$ben, cal$obs_pel, cal$obs_ben)[c(cal$pel, cal$ben, cal$obs_pel, cal$obs_ben) > 0])
plot(y, pos(cal$obs_pel), pch = 16, col = "#1f78b4", log = "y", ylim = yl,
     xlab = "Year", ylab = expression("Catch density (g m"^-2*" yr"^-1*")"), main = "Modelled vs observed catch")
points(y, pos(cal$obs_ben), pch = 16, col = "#b2570b")
lines(y, pos(cal$pel), col = "#1f78b4", lwd = 2); lines(y, pos(cal$ben), col = "#b2570b", lwd = 2)
legend("bottomright", c("pelagic obs", "pelagic model", "benthic obs", "benthic model"),
       pch = c(16, NA, 16, NA), lty = c(NA, 1, NA, 1), lwd = 2,
       col = c("#1f78b4", "#1f78b4", "#b2570b", "#b2570b"), bty = "n", cex = 0.75)

fc <- readRDS(ex("kerguelen_fished_cells.rds"))               # per-cell fished result
pal <- hcl.colors(64, "YlOrRd", rev = TRUE)
plot(fc$lon, fc$lat, pch = 15, cex = 2.1, col = pal[as.integer(cut(log10(pos(fc$catch_tot)), 64))],
     xlab = "Longitude", ylab = "Latitude", main = "Modelled catch (2001-2010)")
points(c(69.2, 73.5), c(-49.3, -53.1), pch = 17)
```

The temporal fit is good in both groups (correlation ≈ 0.77 pelagic, 0.82 benthic):
the model reproduces the catch ramp-up and effort-driven variability, and catch
concentrates over the productive northern plateau where effort and biomass are
highest. Two honest limitations for this region: `q_ben` sits at its upper bound —
the model under-produces the large-bodied benthic (toothfish) catch *level* even at
maximum benthic catchability — and pelagic catch runs somewhat high. Both reflect
that a longline/deep-demersal fishery stresses the size-selective assumptions of a
generic size-spectrum model, not a data problem.

## Going further

* **The full gridded workflow.** For the pre-calibrated large regions the pipeline
  also does per-cell forcing → LME-centering → gridded `q` calibration → fished run;
  see the package's gridded workflow (`GRIDDED_WORKFLOW.md`).
* **Another region.** Swap in any FishMIP regional boundary from
  [Fish-MIP/FishMIP_regions](https://github.com/Fish-MIP/FishMIP_regions), pull its
  forcing, effort and catch from the
  [FishMIP Input Explorer](https://fishmip.global-ecosystem-model.cloud.edu.au/shiny/FishMIP_Input_Explorer/),
  and reuse exactly the `run_cell()` loop and the calibration sketch above.
